---
title: "Using Python from within R"
output: github_document

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  fig.path = "man/figures/README-",
  out.width = "120%",
  out.height = '200%',
  dpi = 500
)
```
## Context
R & Python are two of the most popular languages used in analytics. Each have their unique advantages, and it's now easier than ever before to use them both within the same environment.  

In this notebook, we'll leverage [`reticulate`](https://rstudio.github.io/reticulate/), a package created for using Python from within R.  We'll do this by walking through a common data science use case to provide some intuition for how & when to use this within your own workflows.

From RStudio:

>> The **reticulate** package provides a comprehensive set of tools for interoperability between Python and R. The package includes facilities for:
>>
>>  - Calling Python from R in a variety of ways including R Markdown, sourcing Python scripts, importing Python modules, and using Python interactively within an R session.
>>
>>  - Translation between R and Python objects (for example, between R and Pandas data frames, or between R matrices and NumPy arrays).
>>
>>  - Flexible binding to different versions of Python including virtual environments and Conda environments.

---
## Using reticulate
Let's start by loading the R libraries we'll use for our analysis.
```{r load_r_libs,message=FALSE, warning=FALSE}
library(reticulate) ## load the reticulate library
library(tidyverse) ## loading for data wrangling and visualization
library(glue) ## loading this to help with creating URLs to hit a web API
library(lubridate) ## loading this for help with date operations
```

### Virtual Environments and Python / reticulate
Before we get into an applied example, here's some general housekeeping and helpful info for using `reticulate`:

- Reticulate allows for the use & creation of virtual environments for Python.  
- Virtual environments allow you to define an isolated version of Python that operates completely separate from the main version of Python installed on your computer.  

The packages you install in a virtual environment are isolated as well, which is nice, because otherwise there's a risk you'd overwrite the versions of packages your computer depends on -- which could lead to major issues.  [Here's a link](https://docs.python.org/3/tutorial/venv.html) to virtual environment documentation in case you'd like to learn more about how to leverage this in Python.

This is the way you'd generally initiate a virtual environment for Python in your terminal.
```{bash initiate_py_env, eval=FALSE,message=FALSE,warning=FALSE}
mkvirtualenv r_py3_venv --python=python3 
```

When using `reticulate`, however, the call is slightly different.  Note that I'm using Python 3 in this notebook.
```{r}
virtualenv_create("r_py3_venv", 
                  python = "/usr/local/bin/python3")
use_virtualenv("r_py3_venv")
```

### Using pure Python within RStudio
reticulate also offers some options for using _pure_ Python (i.e. not inclusive of R syntax or operators) as part of the package.

Here's an example where we can let `knitr` run a Python chunk using `{python}` as the first line of the chunk (similar to what you'd do for `R`).
```{python}
print('hello world!')
```

You can also pass a raw string of Python code to the `py_eval` function in Reticulate.
```{r}
## Pull the last element from a Python list
py_eval('[867,"-",5309].pop()')
```


### Installing Python packages
Reticulate also allows for the installation and import of Python packages.  

For reference, here's how you'd generally install Python packages _outside_ of reticulate using the command line and the `pip` package manager (`conda` is also a popular Python package and environment manager, and reticulate has functionality here as well).
```{bash install_py_package1, eval=FALSE}
pip install pandas
pip install requests
```



For installation of packages in `reticulate`, you can specify the virtual environment you want to use and include an R list of the Python packages you'd like to install to the envioronment.  
```{r install_py_package2, eval=FALSE}
py_install(c("pandas",
             "requests",
             "pandas.io"), 
           envname = "r_py3_venv")
```
Note: if you don't specify a virtual environment, reticulate will automatically create one called "r-reticulate" so as not to interfere with your system's default Python installation.

### Importing Python packages
The obvious next step here is to load the libraries we installed so we can use them.  For reference, here's how we'd do this within a Python environment outside of R.
```{python import_py_package1, message=FALSE,eval=FALSE}
import pandas
import requests
import pandas.io.json
```

Now here's the reticulate equivalent.

```{r import_py_package2, message=FALSE}
pandas = import('pandas')
requests = import('requests')
pandas_io = import('pandas.io.json')
```

Once you've imported packages, you can reference the functions within them by using the `<package_name>$<function_name>` notation (as opposed to the `<package_name>.<function_name>` notation in Python.)  

Here's an example with `pandas.pivot()`:
```{r call_pandas_function}
pandas$pivot
```


## Applied Example: scraping weather.gov data
### Context
I'm from Akron, Ohio where [the weather is usually cloudy and it's cold for at least 6 months out of the year](https://www.usclimatedata.com/climate/akron/ohio/united-states/usoh0008).  Living in Northern California is great (for many reasons, but weather is certainly a big one!); I've really been fascinated by the climates of each San Francisco neighborhood and nearby cities.

The National Weather Service (NWS) has a [web API](https://www.weather.gov/documentation/services-web-api) that can be used to retrieve recently-recorded temperatures from weather stations across the US.  To explore the Bay Area's various climates, I decided it'd be fun to pull temperature observations for a list of nearby weather stations & see just how different the temperatures actually are.

To speed up the tutorial here for a moment - I'm going to start with a pre-built list of area weather stations. 
```{r read_station_list, warning=FALSE, message=FALSE}
station_list = read_csv('data/station_identifiers.csv')
```


```{r show_station_list}
station_list
```

### The Weather.gov API
We can hit the web API to grab recent observations from a given station with the following URL structure:
`https://api.weather.gov/stations/<STATION_ID>/observations`



Here's a snippet of the JSON response we get when we hit this API with the San Francisco Airport (station identifier: KSFO) supplied as the weather station.
```{bash show_api_json,message=FALSE}
curl -s https://api.weather.gov/stations/KSFO/observations | tail -n 75
```

### Goals of the Exercise
I can only retrieve data from one station identifier at a time, so I need to build a script that can either loop over these station names in my data and hit this API (or write a function that can be applied over each row of the dataframe).  

Since this data is returned as a nested JSON object, I **also** need to extract each individual temperature observation and place it into a dataframe once I get the initial response.

Web scraping & API interaction are both areas where Python arguably has more robust libraries than R.  While it's certainly not to say there are packages in R that _could_ handle the following work elegantly, I've elected to use Python here for the sake of a simple example.

I think `for` loops are typically easier to explain a process step by step, so I've built one below instead of mapping a funcion over dataframe rows.

Let's start by verbosely breaking out the URL into components that we can glue together as we iterate over weather stations.
```{r build_url_structure}
base_url = 'https://api.weather.gov/stations/'
resource = '/observations?'
station = NULL
```


Ultimately, we want to store each temperature observation from each respective station into a dataframe for us to analyze.  This step creates a 'parking lot' of sorts for our data to be placed in as we go through each part of the loop.
```{r establish_null_dataframe}
temperature_data = data.frame(NULL)
```


Here's where I invoke Python's `requests` and `pandas` libraries to help with fetching & wrangling data from the API.  Have a look at the code for an explanation of each step. 
```{r api_data_wrangling}
## for every row `i` in my dataframe of stations, I want to do the following
for (i in seq_along(station_list$station_identifier)){
  
  ## establish the station identifier I need to use in the API call
  station = station_list$station_identifier[i]
  ## create the URL we'll use to make the API request
  url=glue(base_url,station,resource)
  
  ## use Python's `requests` library to grab the contents of the URL I made from the web
  page=requests$get(url)
  
  ## exclude superfluous information not containing data I care about (i.e. HTPP response, etc)
  content=page$content
  
  ## use `pandas.io` from Python to extract and normalize the JSON from the API request
  content_pandas_io = pandas_io$loads(content)
  
  ## at this point, I have a normalized JSON object stored as a list in R
  ## there is a element in this list for each observation recorded from this station
  ## for every element `j`, I want to extract the time the observation was recorded and the temperature that was recorded (in Fahrenheit)
  for (j in seq_along(content_pandas_io$features)){
    
    ## for each observation, I want a one-row dataframe with columns for date, station, and temperature
    record = tibble(date=content_pandas_io$features[[j]]$properties$timestamp[1],
                        temperature=round(content_pandas_io$features[[j]]$properties$temperature[1]$value*(9/5)+32,2),
                        station_name=station_list$short_name[i],
                        station_id = station)
    
    ## once I have that, I can append this newly-created row to a dataframe containing the final results
    temperature_data=bind_rows(record,temperature_data)
  }
  
}

```


These API calls will occasionally return observations from further back in time, so let's filter them out and convert the UTC timestamps to local time.
```{r filter_data_convert_date, message=FALSE}
temperature_data = 
  temperature_data %>% 
  filter(as.Date(date)>=as.Date(today()-days(2))) %>% 
  mutate(observation_date_time_local=as_datetime(date,tz='America/Los_Angeles'))
```

### Visualizing Results
Now that we have recent data from all weather stations, converted temperatures from Celsius to Farenheit, and converted timezones...let's visualize and see what we learn!  Here's a plot created with ggplot2 that shows the fluctuation of temperatures at each of these stations' observations from the past few days.
```{r build_plot}
ggplot(data=temperature_data,
       aes(x=observation_date_time_local,
           y=temperature)) + 
  geom_line(color='gray') + 
  scale_color_gradient(low='blue',
                       high='orange') +
  geom_point(aes(color=temperature)) + 
  facet_wrap(~station_name)
```

## Session Info
```{r session_info, include=TRUE, echo=TRUE, results='markup'}
sessionInfo()
```